---
title: "Consistency of Random Forest/Randomer Forest"
author: "Jaewon Chung"
output: 
  html_document:
    code_folding: hide
    fig_caption: yes
    fig_height: 8
    fig_width: 8
    highlight: pygments
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 2
  keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

```{r, eval = FALSE, include = FALSE, echo = FALSE}
require(devtools)
rmarkdown::render("Consistency.Rmd")
system('open Consistency.html')
```

```{r packages, include = FALSE, echo = FALSE}
library('ggplot2')
library('rerf') # reproducible only with CRAN v 1.1.3 or set max.depth to log2(n)
library('cowplot')
library('gridExtra')
library('grid')
```

# Consistency of Random Forest/Randomer Forest

[Biau et al](http://www.jmlr.org/papers/volume9/biau08a/biau08a.pdf) (2016) proposed a distribution for which Breiman's RF is inconsistent. The joint distribution of $(X, Y)$ is as follows:
$X$ has a uniform distribution on $[0,1] \times [0,1] \cup [1,2] \times [1,2] \cup [2,3] \times [2, 3]$. $Y$ is deterministic function of $X$, that is $f(X) \in \{0, 1\}$. The $[0,1] \times [0, 1]$ square is divided into countably infinite vertical stripes, and $[2, 3] \times [2,3]$ square is similarly divided into countably infinite horizontal stripes. In both squares, the stripes with $f(X) = 0$ and $f(X) = 1$ alternate. The $[1, 2] \times [1, 2]$ square is a $2 \times 2$ checker board. The joint distribution is shown below.

## Joint Distribution of $(X, Y)$
```{r, fig.width = 6, fig.height = 6}
getY <- function(n) {
  sequence <- 1/(2^seq(1, n))
  sequence <- c(0, cumsum(rep(sequence / 2, each = 2)))
  
  return(sequence)
}

getYFlip <- function(n) {
  sequence <- 1/(2^seq(1, n))
  sequence <- c(0, cumsum(rev(rep(sequence / 2, each = 2))))
  
  return(sequence)
}

drawDistribution <- function(n) {
  sequence <- getY(n)
  sequence.flip <- getYFlip(n)
  colors <- rep(c('white', 'darkgray'), (length(sequence) - 1) / 2)
  
  len <- length(sequence)
  
  # Df for lower left square
  df <- data.frame(xmin = sequence[1:len - 1], 
                 xmax = sequence[2:len], 
                 ymin = rep(0, (len - 1)),
                 ymax = rep(1, (len - 1)),
                 fill = colors
  )
  
  # Df for middle square
  df2 <- data.frame(xmin = c(1, 1, 1.5, 1.5), 
                    xmax = c(1.5, 1.5, 2, 2), 
                    ymin = c(1, 1.5, 1, 1.5), 
                    ymax = c(1.5, 2, 1.5, 2),
                    fill = c('white', 'darkgray', 'darkgray', 'white'))
  
  # Df for top right square
  df3 <- data.frame(xmin = rep(2, (len - 1)),
                    xmax = rep(3, (len - 1)),
                    ymin = sequence.flip[1:len - 1] + 2,
                    ymax = sequence.flip[2:len] + 2,
                    fill = rev(colors))
  
  df <- rbind(df, df2, df3)

  fig <- ggplot(df) + 
    geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill))

  fig <- fig + geom_rect(aes(xmin = 1, xmax = 2, ymin = 1, ymax = 2), alpha = 0, color = 'black') +
        geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1), alpha = 0, color = 'black') +
        geom_rect(aes(xmin = 2, xmax = 3, ymin = 2, ymax = 3), alpha = 0, color = 'black')

  fig <- fig + scale_fill_identity() + coord_equal() + theme_gray(base_size = 20) +
    #theme(panel.background = element_blank()) + 
    #labs(title = "Distribution of Y") +
    scale_y_continuous(expand = c(0,0)) + scale_x_continuous(expand = c(0,0)) + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    panel.background = element_blank(), axis.line = element_line(colour = "black", size = 1)) + 
    labs(x = expression(X[1]), y = expression(X[2]))

  return(fig)
}

joint.distribution <- drawDistribution(10)
joint.distribution
```

## Sample Draw (n=1000)
```{r, fig.width = 6, fig.height = 6}
createData <- function(n, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  # Generate distribution of Y
  sequence <- getY(50)
  sequence.flip <- getYFlip(50)
  df <- data.frame(min = sequence[1:length(sequence) - 1], 
                   max = sequence[2:length(sequence)], 
                   label = rep(c(0, 1), (length(sequence) - 1) / 2))
  
  df2 <- data.frame(min = sequence.flip[1:length(sequence.flip) - 1] + 2, 
                    max = sequence.flip[2:length(sequence.flip)] + 2, 
                    label = rep(c(1, 0), (length(sequence.flip) - 1) / 2))
  
  df <- rbind(df, df2)
  
  squares <- sample(c(1, 2, 3), n, replace = TRUE)
  Y <- integer(n)
  X <- matrix(0, nrow = n, ncol = 2)
  
  for (i in 1:n) {
    square <- squares[i]
    point <- runif(2, min = square - 1, max = square)
    
    X[i, ] <- point
    if (square == 2) {
      if ((point[1] <= 1.5 & point[2] > 1.5) | (point[1] > 1.5 & point[2] <= 1.5)) {
        Y[i] <- 1
      }
    } else if (square == 1) {
      Y[i] <- df$label[(point[1] >= df$min) & (point[1] < df$max)]
    } else if (square == 3) {
      Y[i] <- df$label[(point[2] >= df$min) & (point[2] < df$max)]
    }
  }
  return(list(X = X, Y = factor(Y)))
}

visualizeData <- function(n) {
  out <- createData(n, seed = NULL)
  X <- out$X
  Y <- out$Y
  
  qplot(X[, 1], X[, 2], color = Y) + labs(x = expression(X[1]), y = expression(X[2]),  title = paste0("Visualization of sample data (n=", n, ")"))
}

visualizeData(1000)
```

# Running Experiment
Error rates are estimated by taking a random sample of size $n$, training the classifiers (RF/RerF), and computing the fraction misclassified in a test set of 10,000 samples. This is repeated ten times for each value of $n$. The reported error rate is the mean over the ten repeated experiments.

```{r, echo=FALSE, eval=FALSE, results='hide'}
runExperiment <- function(n.samples, seed = NULL) {
  errors.rerf <- integer(length(n.samples))
  errors.rf <- integer(length(n.samples))
  
  for (i in 1:length(n.samples)) {
    res <- createData(n.samples[i], seed = seed)
    X <- res$X
    Y <- res$Y
    
    res2 <- createData(10000, seed = seed)
    X.test <- res2$X
    Y.test <- res2$Y
    
    p <- ncol(X)
    d <- ceiling(sqrt(p))
    rho <- 1 / p
    prob <- 0.5
    
    mat.options <- list(p = p, d = d, random.matrix = 'binary', rho = rho, prob = prob)
    
    print(paste("Running forest with rerf", '(n =', n.samples[i], "samples)"))
  
    forest <- RerF(X, Y, mat.options = mat.options, store.oob = T, seed = seed, num.cores = 1)
    Yhat <- Predict(X.test, forest, num.cores = 1)
    errors.rerf[i] <- sum(Y.test != Yhat) / length(Yhat)
    
    mat.options <- list(p = p, d = d, random.matrix = 'rf', rho = rho, prob = prob)
    
    print(paste("Running forest with rf", '(n =', n.samples[i], "samples)"))
  
    forest <- RerF(X, Y, mat.options = mat.options, store.oob = T, seed = seed, num.cores = 1)
    Yhat <- Predict(X.test, forest, num.cores = 1)
    errors.rf[i] <- sum(Y.test != Yhat) / length(Yhat)
  }
  return(list(errors.rerf = errors.rerf, errors.rf = errors.rf))
}
errors <- vector('list', 10)
n.samples <- c(100, 1000, 5000, 10000)
for (i in 1:10) {
  print(paste0('Run ', i))
  errors[[i]] <- runExperiment(n.samples = n.samples, seed = 20181 + i)
}
```

```{r, echo=TRUE}
# Data wrangling for plotting
errors.rerf <- matrix(NA, ncol = 4, nrow = 10)
errors.rf <- matrix(NA, ncol = 4, nrow = 10)
for (i in 1:10) {
  tmp <- errors[[i]]
  errors.rerf[i, ] <- tmp$errors.rerf
  errors.rf[i, ] <- tmp$errors.rf
}
errors.rerf.se <- apply(errors.rerf, 2, sd)
errors.rf.se <- apply(errors.rf, 2, sd)
errors.rerf.se <- errors.rerf.se / sqrt(length(errors.rerf.se))
errors.rf.se <- errors.rf.se / sqrt(length(errors.rf.se))
errors.rf <- colMeans(errors.rf)
errors.rerf <- colMeans(errors.rerf)
df <- data.frame(rep(n.samples, 2), c(errors.rerf, errors.rf), c(errors.rerf.se, errors.rf.se), c(rep('RerF', length(n.samples)), rep('RF', length(n.samples))))
colnames(df) <- c('n.samples', 'errors', 'se', 'Algorithm')
```

## Plotting Results
Results show that RF cannot achieve an error rate lower than 1/6. This is because RF will always choose to split either in the lower left square or top right square and never in the center square. On the otherhand, RerF achieves close to zero error.

```{r, echo=TRUE, fig.height=6, fig.width=8}
err.plot <- ggplot(data = df, aes(x = n.samples, y = errors, color = Algorithm)) + 
  geom_line(size=1) +  
  geom_errorbar(aes(ymin=errors-se, ymax=errors+se), size = .9) +
  geom_hline(aes(yintercept = 1/6, linetype='g'), color = 'black', show.legend=FALSE, alpha = .8) +
  scale_linetype_manual(values = 2) +
  scale_x_continuous(trans='log10', breaks = c(100, 1000, 5000, 10000), labels = c(.1, 1, 5, 10)) + 
  scale_y_continuous(expand = c(0,0)) +
  scale_color_manual(values=c("red", "black")) +
  labs(x = "n (in thousands)", y = "Error Rate")
err.plot <- err.plot +
  theme_gray(base_size = 20) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    panel.background = element_blank(), axis.line = element_line(colour = "black", size = 1), aspect.ratio=1)
# Get legend
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}
legend <- g_legend(err.plot)
out <- plot_grid(joint.distribution, err.plot + theme(legend.position = 'none'), legend, labels = c("(A)", "(B)"), ncol = 3, rel_widths = c(3.7, 4, 1.5), label_size = 20)
out
```

```{r, echo=TRUE, eval=FALSE}
ggsave('error_rate_consistency.pdf', plot = out, width = 7, height = 4)
```
